Storm Data Analysis - Week 4

### 2021-03-23 Manuscript.Rmd
Completed element selection; started factor (event type) aggregation

# Min-max normalization
normalize <- function(x) {
    return ((x - min(x)) / (max(x) - min(x)))
  }

@@

```{r exponents}
unique(data$PROPDMGEXP)
unique(data$CROPDMGEXP)
```

@@

# Clean up
rm(temp_data)
```

@@

# Select useful columns
lean_data <- mdata %>%
    select(REFNUM, event, year, c_timezone,
           FATALITIES, INJURIES, tot_dmg)
# Economical damages aren't always available, so better have a tibble with all
# complete cases
dmg_data <- lean_data %>%
    filter(!is.na(tot_dmg))

@@

We will now plot the distribution of values of interests, so as to only select significant *event types* in terms of occurrence. We'll work with the 90^th^ percentile (and above) of event frequency.

@@

# Make a copy of data and cast as DT
dt_data <- as.data.table(lean_data)
# Define the number of rows that will be in the grouped DT. This is equal to the
# unique values of event, as we'll group by that.
denominator <- uniqueN(dt_data, "event")
# Apply final transformations to DT
sel_data <- dt_data[, .(
                        counter = .N, # Count elements in group
                        # Total monetary damages per group
                        total_dmg = sum(tot_dmg, na.rm = TRUE),
                        # Total fatalities
                        fatalities = sum(FATALITIES, na.rm = TRUE),
                        # Injuries
                        injuries = sum(INJURIES, na.rm = TRUE)
                        ),
                    # Grouping by event type
                    by = event][
                        ,
                        # Compute a "percent rank": rank events by number of
                        # occurrences (desc), then assign a percentile by
                        # dividing rank by denominator (i.e. nrows)
                        # Note double dot syntax to access calling scope
                        `:=`(perc_rank = frankv(counter)/..denominator,
                             # Same for monetary damage
                             dmg_rank = frankv(total_dmg)/..denominator,
                             # And life damage
                             inj_rank = frankv(injuries)/..denominator,
                             fat_rank = frankv(fatalities)/..denominator
                             )
                        ][
                            # Filter for frequency rank >=95th
                            perc_rank >= 0.90
                        ][
                            order(-counter)
                        ]

@@

Figure 1 is a panel plot of data at or above the 95^th^ percentile of (A) event frequency by type, (B) monetary damages and (C) damages to persons in the period of interest (1950–2013). One can easily that the magnitude of event-related damages decreases almost exponentially with frequency. It should be noted that the preponderant majority of events below 95^th^ percentile of frequency are actually data-entry errors arising from "event type" being an open-text field in the database.

```{r frequency-plots}

sel_data %>% ggplot(
  aes(x = fct_reorder(event, perc_rank, .desc = T), y = counter)
) +
  geom_point() +
  scale_y_continuous(
    # limits = c(0, 100), 
    name = ("Frequency quantile\n")) +
  theme_bw()

>>>>

# Min-max normalization
normalizer <- function(x) {
    return ((x - min(x)) / (max(x) - min(x)))
  }

@@

```{r exponents}
unique(fdata$PROPDMGEXP)
unique(fdata$CROPDMGEXP)
```

@@

# Clean up
# rm(temp_data)
```

@@

# Select useful columns
lean_data <- mdata %>%
    select(REFNUM, event, year, c_timezone,
           FATALITIES, INJURIES, tot_dmg)

@@

@@

# Make a copy of data and cast as DT
sel_data <- as.data.table(lean_data)
# Define the number of rows that will be in the grouped DT. This is equal to the
# unique values of event, as we'll group by that.
denominator <- uniqueN(dt_data, "event")
# Make a copy for future use:
tot_data <- copy(sel_data)
# Apply final transformations to DT
sel_data[, .(
    counter = .N, # Count elements in group
    # Total monetary damages per group
    total_dmg = sum(tot_dmg, na.rm = TRUE),
    # Total fatalities
    fatalities = sum(FATALITIES, na.rm = TRUE),
    # Injuries
    injuries = sum(INJURIES, na.rm = TRUE)
  ),
  # Grouping by event type
  by = event][
    ,
    # Compute a "percent rank": rank events by number of
    # occurrences (desc), then assign a percentile by
    # dividing rank by denominator (i.e. nrows)
    # Note double dot syntax to access calling scope
    `:=`(perc_rank = frankv(counter)/..denominator,
         # Same for monetary damage
         dmg_rank = frankv(total_dmg)/..denominator,
         # And life damage
         inj_rank = frankv(injuries)/..denominator,
         fat_rank = frankv(fatalities)/..denominator
    )
  ][
    # Filter for frequency rank >=95th
    perc_rank >= 0.90
  ][
    order(-counter)
  ]

@@

```{r ev-selection-comparison}

gdata <- as.data.table(temp_data)
# With DT syntax, group and compute relevant sums
totals <- gdata[,.(sum_inj = sum(INJURIES, na.rm = T),
               sum_fat = sum(FATALITIES, na.rm = T),
               sum_dmg = sum(tot_dmg, na.rm = T))]

selected <- sel_data[,.(sum_inj = sum(INJURIES, na.rm = T),
               sum_fat = sum(FATALITIES, na.rm = T),
               sum_dmg = sum(tot_dmg, na.rm = T))]

totals == selected
```

Events below the 90^th^ percentile of frequency are mostly irrelevant to the analysis. To confirm this, we calculated total damages, injuries and fatalities from the global data and from the selected data. Results are identical (see above), despite the latter table having only a fraction of the event types (`r length(unique(gdata$event))` _vs._ `r length(unique(sel_data$event))`).

### Aggregation of events by type

```{r event-aggregation}

# Functions to aggregate (collapse) factor levels require specific values, specified in a named vector. Here, we use "data mining" to generate such vector.

# Create regexp patterns to match to. The following is also the priority order
# with which events will be assigned to a category.
stormy <- c("storm*", "thunder*", "tstm", "rain", "precip")
maritime <- c("sea*", "ocean*", "mari[nt]*", "curr*", "tid*", "surf*")
